A Wait-free Queue with Poly-logarithmic Step Complexity

### 2022-12-15 podc_paper/related.tex
finished editing introduction

%%
%% This is file generated with the docstrip utility from the source file: 
%%
%% samples.dtx  (with options: `acmsmall-conf')
%% With some modifications for PODC 2023.
%% For the copyright see the source file.
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall,nonacm,anonymous]{acmart}

\usepackage{tikz-qtree}
\usepackage{algorithm}
\usepackage{algpseudocode}
\makeatletter
\renewcommand{\ALG@beginalgorithmic}{\footnotesize}
\makeatother
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{relsize}
\usepackage{enumitem}
\usepackage{bold-extra}
\renewcommand*\contentsname{Table of Contents}

\usepackage{multicol}
\setlength\columnsep{24pt}

\algnewcommand\algorithmicforeach{\bf{for each}}
\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}

\algdef{S}[FUNCTION]{Function}
   [3]{{\tt{\sl{#1}}} {\tt{#2}}\ifthenelse{\equal{#3}{}}{}{\tt{(#3)}}}

\algdef{E}[FUNCTION]{EndFunction}
   [1]{\algorithmicend\ \tt{{#1}}}

\algrenewcommand\Call[2]{\tt{{#1}\ifthenelse{\equal{#2}{}}{}{(#2)}}}

\newcommand\keywordfont{\sffamily\bfseries}
\algrenewcommand\algorithmicend{{\keywordfont end}}
\algrenewcommand\algorithmicfor{{\keywordfont for}}
\algrenewcommand\algorithmicforeach{{\keywordfont for each}}
\algrenewcommand\algorithmicdo{{\keywordfont do}}
\algrenewcommand\algorithmicuntil{{\keywordfont until}}
\algrenewcommand\algorithmicfunction{{\keywordfont function}}
\algrenewcommand\algorithmicif{{\keywordfont if}}
\algrenewcommand\algorithmicthen{{\keywordfont then}}
\algrenewcommand\algorithmicelse{{\keywordfont else}}
\algrenewcommand\algorithmicreturn{{\keywordfont return}}

\renewcommand\thealgorithm{}
\newcommand{\setalglineno}[1]{%
  \setcounter{ALC@line}{\numexpr#1-1}}

\newcommand{\sub}[1]{\textsubscript{#1}}
\renewcommand{\tt}[1]{\texttt{#1}}
\renewcommand{\sl}[1]{\textsl{#1}}
\renewcommand{\it}[1]{\textit{#1}}
\renewcommand{\sc}[1]{\textsc{#1}}
\renewcommand{\bf}[1]{\textbf{#1}}
\newcommand{\nf}[1]{{\normalfont{\texttt{#1}}}}
\newcommand{\cmt}[1]{\Comment{#1}}
\newcommand{\head}{head}
\newcommand{\size}{size }

\usepackage{amsmath,amssymb,amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{observation}[theorem]{Observation}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{invariant}[theorem]{Invariant}
\newtheorem{proposition}[theorem]{Proposition}

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2023}
\acmYear{2023}
\acmDOI{XXXXXXX.XXXXXXX}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[PODC 2023]{ACM Symposium on Principles of Distributed Computing}{June 19--23,
  2023}{Orlando, FL}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
\acmPrice{15.00}
\acmISBN{978-1-4503-XXXX-X/18/06}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{A Wait-free Queue with Poly-logarithmic Worst-caseStep Complexity}
%\title{The Name of the Title Is Hope}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Eric Ruppert}
\affiliation{%
  \institution{York University}
  \streetaddress{P.O. Box 1212}
  \city{Toronto}
  \state{Ontario}
  \country{Canada}
  \postcode{43017-6221}
}
\authornote{Both authors contributed equally to this research.}
\email{ruppert@cse.yorku.ca}
\orcid{1234-5678-9012}

\author{Hossein Naderibeni}
\authornotemark[1]
\email{hnaderi268@gmail.com}
\affiliation{%
  \institution{York University}
  \streetaddress{P.O. Box 1212}
  \city{Toronto}
  \state{Ontario}
  \country{Canada}
  \postcode{43017-6221}
}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{Trovato et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
In this work, we introduce a novel linearizable wait-free queue implementation.
Linearizability and lock-freedom are standard requirements for designing shared data structures. To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case, called the CAS Retry Problem. We show that our algorithm avoids this problem with the helping mechanism which we use and has a worst-case running time better than prior lock-free queues. The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010520.10010553.10010562</concept_id>
  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010575.10010755</concept_id>
  <concept_desc>Computer systems organization~Redundancy</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010553.10010554</concept_id>
  <concept_desc>Computer systems organization~Robotics</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>10003033.10003083.10003095</concept_id>
  <concept_desc>Networks~Network reliability</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Computer systems organization~Embedded systems}
\ccsdesc[300]{Computer systems organization~Redundancy}
\ccsdesc{Computer systems organization~Robotics}
\ccsdesc[100]{Networks~Network reliability}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{concurrent data structures, wait-free queues}

\received{20 February 2007}
\received[revised]{12 March 2009}
\received[accepted]{5 June 2009}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. It is not hard to see why multiple processes cannot share a sequential data structure. For example, consider two processes trying to append to a sequential linked list simultaneously. Processes $P,Q$ read the same tail node, $P$ changes the next pointer of the tail node to its new node, and then $Q$ does the same. In this run, $P$'s update is overwritten, which makes the queue inconsistent. One solution is to use locks; whenever a process wants to update or query a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only the process holding the lock can progress when the data structure is locked. For these reasons, we are interested in the design of an efficient shared queue without using locks.

A sequential queue stores a sequence of elements and supports two operations, enqueue and dequeue. \texttt{Enqueue(e)} appends the element \texttt{e} to the sequence stored. \texttt{Dequeue} removes and returns the first element in the queue. If the queue is empty, it returns \texttt{null}. In a concurrent version of the queue data structure, operations do not happen one at a time. The question that may arise is, ``What properties matter for the implementation of a shared data structure?'' Since executions on a shared data structure are different from sequential ones, the correctness conditions also differ. To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions. A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.

A system is called \it{asynchronous} when processes in the system run at arbitrarily varying speeds, i.e., the scheduling of each process is independent from the scheduling of other processes. Our model is an asynchronous shared-memory distributed system of $p$ processes. Herlihy \cite{10.1145/114005.102808} showed that one cannot implement concurrent versions of all data structures with multi-reader multi-writer registers alone and introduced a hierarchy based on how powerful objects are to solve the consensus problem among processes. Objects like LL/SC or CAS can be used to reach a consensus among any number of processors. A CAS object provides an atomic Compare\&Swap(\nf{new, old}) operation: if the value stored in the object is \nf{old}, it updates the value to \nf{new} and returns \nf{true}; otherwise it returns \nf{false}.  In this work, we use Compare\&Swap (\nf{CAS}) objects to synchronize among processes.

% \bf{Concurrent execution: I didn't understed interleaving steps of processes}

The standard safety condition is called \textit{linearizability}~\cite{10.1145/78969.78972}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response.

There are various progress properties; the strongest is wait-freedom, and the more common is lock-freedom. An algorithm is \textit{wait-free} if each operation terminates after a finite number of its own steps. We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates. A wait-free algorithm is also lock-free but not vice versa; in an infinite run of a lock-free algorithm, there might be an operation that takes infinitely many steps but never terminates.

We will present a wait-free linearizable queue, which to the best of our knowledge  is the first wait-free queue whose operations run in a poly-logarithmic number of steps. We design a tournament tree in which each process tries to propagate its operations along a path from the process's leaf up to the root to be linearized. Processes can do queries to get information about the linearization stored in the root. We create a helping mechanism for when a process is propagating its operation to a node, such that new operations are ensured to be propagated one step further up the tree after at most 2 \nf{CAS} invocations. The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized. \nf{CAS} instructions cost more than other instructions for the processor. Each operation in our queue does $\Theta(\log p)$ \nf{CAS} operations compared to $\Omega(p)$ \nf{CAS} steps for previous queues in the worst case. We use unbounded memory space in our queue and present a way to reduce the memory size to the total number of operations. Furthermore, we do not address garbage collection but we describe a way we think it can be handled.

The rest of the thesis is organized as follows. Chapter 2 gives an outline of the related work done in the area and the motivation for our implementation. Previous lock-free queues and their common problem are presented in Section 2.1. In Section 2.2, we mention some restricted lock-free queues. Section 2.3 talks about poly-logarithmic constructions of shared objects. Section 2.4 ends with a lower bound on the amortized time complexity of shared queues.

In Chapter 3 we introduce a poly-logarithmic step wait-free queue. In Section 3.1 we give a high-level description of our implementation. We also discuss the motivation and requirements in our design to achieve poly-log time. Section 3.2 contains the algorithm itself and there is an example of a concurrent execution of the algorithm described in Section 3.3.

We prove the correctness of our queue in Chapter 4 by showing it is linearizable.

Chapter 5 is devoted to a complexity analysis for our implementation. We compute the number of \nf{CAS} instructions our algorithm invokes and the worst-case and amortized running time of the queue. In the end, we prove our queue is wait-free.

Finally, we give some concluding remarks in Chapter 6. It contains how we can improve our algorithm's memory usage and how to use its idea in designing other wait-free data structures.

@@

\section{Queue Implementation} \label{DescriptQ}

In our model there are $p$ processes doing \nf{Enqueue} and \nf{Dequeue} operations on a queue concurrently. We design a linearizable wait-free queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the number of elements in the queue at the time of linearization. We avoid the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays, which suffer from the CAS Retry Problem.

There is a shared binary tree among the processes to agree on one total ordering of the operations invoked by processes. The tree is called a \it{tournament tree} (see Figure~\ref{fig::blocktree}). Each process has a leaf in which the operations invoked by the process are stored in order. When a process wishes to do an operation it appends the operation to its leaf and tries to propagate its new operation up to the tree's root. Each node of the tree keeps an ordering of operations propagated up to it. All processes agree on the sequence of operations in the root and this ordering is used as the linearization ordering. 
\begin{figure}[hbtp]
\begin{center}
\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
\end{center}
\caption[Tournament tree.]{\label{fig::blocktree}Each of the processes $P_1,P_2,...,P_p$ has a leaf and in each node there is an ordering of operations stored. Each process tries to propagate its operations up to the root, which stores a total ordering of all operations.}  
\end{figure}

To propagate operations to a node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, merges them to create an ordering and then tries to append the ordering to the sequence stored in $n$. We call this procedure \tt{$n$.Refresh} (see Figure \ref{fig::propagstep}). A \nf{Refresh} on $n$ with a successful append helps to propagate their operations up to $n$. We shall prove that if a process invokes \tt{Refresh} on the node $n$ two times and fails to append the new operations to $n$ both times, the operations that were in $n$'s children before the first \tt{Refresh} are guaranteed to be in $n$ after the second failed \tt{Refresh}.
We sketch the argument here.

\begin{figure}[hbtp]
\begin{center}
\hspace{6em}\begin{subfigure}[b]{.29\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
\Tree [.$r_1,l_1,l_2,r_2,l_3$
$l_1,l_2,l_3,l_4,l_5$
$r_1,r_2,r_3,r_4$ ]
\end{tikzpicture}}
  \caption{Before the \nf{Refresh}.}
\end{subfigure}
\hfill
\begin{subfigure}[b]{.29\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
\Tree [.$r_1,l_1,l_2,r_2,l_3,l_4,l_5,r_3,r_4$
$l_1,l_2,l_3,l_4,l_5$
$r_1,r_2,r_3,r_4$ ]
\end{tikzpicture}}
  \caption{After the \nf{Refresh}.}
\end{subfigure}
\hspace{6em}
\caption[A successful \nf{Refresh}.]{\label{fig::propagstep} Before and after a \nf{$n$.Refresh} with a successful append. Operations propagating from the left child are labelled with $l$ and from the right child with $r$.}
\end{center}
\end{figure}

\begin{figure}[hbpt]
  \center\includegraphics[width=4in]{pics/doublyrefresh.drawio.png}
  \caption[Two consecutive failed \nf{Refresh}es by a process.]{\label{fig::simpleDoubleRefresh}Time relations between the concurrent successful \nf{Refresh}es and the two consecutive failed \nf{Refresh}es.}
\end{figure}

We use \nf{CAS} (Compare\&Swap) instructions to implement the \nf{Refresh}'s attempt to append  described in the previous paragraph.
The second failed \nf{Refresh} of $P$ is assuredly concurrent with a successful \texttt{Refresh} that has read its information after the invocation of the first failed \nf{Refresh} (see Figure \ref{fig::simpleDoubleRefresh}). This is because some process $L$ does a successful append during $P$'s first failed attempt, and some process $K$ performs a \nf{Refresh} that reads its information after $L$'s append and then performs a successful append during $P$'s second failed \nf{Refresh}. Process $K$'s \nf{Refresh} helps to append the new operations that were in $n$'s children before $P$'s first failed \nf{Refresh}, in case they were not already appended. After a process appends its operation into its leaf it can call \nf{Refresh} on the path up to the root at most two times on each node. So, with $O(\log p)$ \nf{CAS}es an operation can ensure it appears in the linearization. This cooperative solution allows us to overcome the CAS Retry Problem.

It is not efficient to explicitly store the sequence of operations in each node because each operation would have to be copied all the way up to the root; doing this would not be possible in poly-logarithmic time.
Instead we use an implicit representation of the operations propagated together. Furthermore, we do not need to maintain an ordering on operations propagated together in a node until they have reached the root. It is sufficient to only keep track of sets of operations propagated together in each \nf{Refresh} and then define the linearization ordering only in the root (see Figure \ref{fig::set}). Achieving a constant-sized implicit representation of operations in a \nf{Refresh} allows us to do \texttt{CAS} instructions on fixed-size objects in each \tt{Refresh}. To do that, we introduce \it{block}s. A block stores information about the operations propagated by a \nf{Refresh}. It contains the number of operations from the left and the right child propagated to the node by the \texttt{Refresh} procedure. See Figure~\ref{fig:block} for an example. A node stores an array of blocks of operations propagated up to it. A propagate step  aggregates the new blocks in the children into a new block and stores the new block in the parent. We call the aggregated blocks \it{subblocks} of the new block and the new block the \it{superblock} of them.  In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one process cannot invoke two operations concurrently. Thus, there are at most $p$ operations in a block. Furthermore, since the operations in a \texttt{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together. Our choice is to put the operations propagated from the left child before the operations propagated from the right child. In this way, if we know the number of operations from the left child and the number of operations from the right child in a block, we have a complete ordering of the operations.

\begin{figure}[hbtp]
\begin{center}
\begin{tikzpicture} [level 1/.style={level distance=1.8cm,sibling distance=0.6cm}, level 2/.style={level distance=1.4cm,sibling distance=0.4cm}]

\Tree [.{$\big\{op_2^1,op_1^1,op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_2^2, op_4^2\big\},\big\{op_1^2\big\}...$ }  [.{ $\big\{op_2^1, op_1^1\big\},\big\{op_2^2\big\},\big\{op_1^2\big\}...$ }
      {\begin{tabular}{|l|c|c|c}  \hline $op_1^1$ & $op_1^2$ & ... \\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c}  \hline $op_2^1$ & $op_2^2$ & ... \\ \hline \end{tabular}} ] [.{ $\big\{op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_4^2\big\},...$ } {\begin{tabular}{|l|c|c|c}  \hline $op_3^1$ & $op_3^2$ & ... \\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c} \hline $op_4^1$ & $op_4^2$ & ... \\ \hline\end{tabular}} ] ]
\end{tikzpicture}
\caption[The operations merged in a \nf{Refresh} step with sets.]{\label{fig::set} Leaves are for processes $P_1$ to $P_4$ from left to right. In each internal node one can arbitrarily linearize the sets of concurrent operations  propagated together in a \nf{Refresh}. For example $op_4^1$ and $op_3^2$ have propagated  together in one \nf{Propagate} step and they will be propagated up to the root together. Since their execution time intervals overlap, they can be linearized in any order.}
\end{center}
\end{figure}

\begin{figure}[hbtp]
\begin{center}
\begin{tikzpicture}
[level 1/.style={level distance=3.3cm,sibling distance=0cm},
	level 2/.style={level distance=2.5cm,sibling distance=0cm},
	level 3/.style={level distance=1.7cm,sibling distance=2cm}]
\begin{small}

\Tree [.{\begin{tabular}{|c|c|c|}  \hline (0,11) & (12,6) & (8,36) \\ \hline\end{tabular}}
 [.{\begin{tabular}{||c||c||}  \hline (10,2) & (5,3) \\\hline  \end{tabular}}
 [.{\begin{tabular}{||c|c||c||}\hline (2,3) & (4,1) & (0,5) \\\hline\end{tabular}} $\vdots$ $\vdots$ ]
  [.{\begin{tabular}{||c||c||}  \hline (0,2) & (1,2)  \\ \hline  \end{tabular}} $\vdots$ $\vdots$ ] ]
          [.{\begin{tabular}{||c||c||c|c||}  \hline (3,8) & (6,0) & (14,6) & (0,16) \\ \hline \end{tabular}}
           [.{\begin{tabular}{||c||c||c|c||}  \hline (0,3) & (4,2) & (5,1) & (3,5) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ]
            [.{\begin{tabular}{||c|c||c||c||}  \hline (2,1) & (5,0) & (4,2) & (9,7) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ] ] ]
\end{small}
\end{tikzpicture}
\end{center}
\caption[The operations merged in a \nf{Refresh} step with (left, right) blocks.]{\label{fig:block} Using blocks to represent operations. Blocks between two lines $||$ are propagated together to the parent. Each block consists of a pair (left, right) indicating the number of operations from the left and the right child, respectively. For example, (12,6) in the root contains  (10,2) from the left child and (6,0) from the right child. The third block in the root (8,36) is created by merging (5,3) from the left child and (14,6) and (0,16) from the right child. (5,3) is the superblock of  (0,5) and (1,2) and (5,1),(3,5) and (4,2) are subblocks of (14,6).}
\end{figure}

So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root. With this agreement on the linearization ordering, we can design a universal construction; for a given object, we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then returning the response for $op$. However, this approach is not enough for an efficient queue.
We show that we can build an efficient queue if we can compute two things about the ordering in the root: (1) the $i$th propagated operation and (2) the rank of a propagated operation in the linearization. We explain how to implement (1) and (2) in poly-logarithmic steps.

After propagating an operation \textit{op} to the root, processes can find out information about the linearization ordering using (1) and (2). 
To get the $i$th operation in the root, we find the block $B$ containing the $i$th operation in the root, and then recursively find the subblock of $B$ in the descendent of the root that contains that $i$th operation. When we reach a block in a leaf, the operation is explicitly stored there. To make this search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search for the required block. We also store pointers to determine the range of subblocks of a block to make the binary search faster. In each block, we store the prefix sum of operations from the left child and from the right child. Moreover, for each block, we store two attributes \nf{end\sub{left}} and \nf{end\sub{right}}, the indices of the last left and right subblock (see Figure \ref{fig::pointer}). We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the \nf{end\sub{left}} and \nf{end\sub{right}} indices of a block and its previous block reduce the search range size to \textsc{O}$(p)$.

To compute the rank in the root of an operation in a leaf, we need to find the superblock of the block that the operation is in. After a block is installed in a node we store the approximate index of its superblock in it to make this faster.

\begin{figure}[hbtp]
\centering
  \includegraphics[width=4in, height=1.4in]{pics/pointers}
  \caption{Each block stores the index of its last subblock in each child. \label{fig::pointer}}
\end{figure}

In an execution on a queue where no dequeue operation returns \nf{null}, the $k$th dequeue returns the argument of the $k$th enqueue. In the general case a dequeue returns \nf{null} if and only if the queue size after the previous operation is 0. We refer to such a dequeue as a \it{null dequeue}. If the dequeue is the $k$th non-null dequeue, it returns the argument of the $k$th enqueue. Having the size of the queue after an operation we can compute the number of non-null dequeues from the number of enqueues before the operation. So, if we store the size of the queue after each block of operations in the root, we can compute the index of the enqueue whose argument is the response to a given dequeue in constant time.

In our case of implementing a queue, a process only needs to compute the rank of a \nf{Dequeue} and get an \nf{Enqueue} with a specific position.  We know we can linearize operations in a block in any order; here, we choose to put \nf{Enqueue} operations in a block before \nf{Dequeue} operations. Consider the following operations, where operations in a cell are concurrent.
\begin{table}[hbtp]
\centering
\begin{small}
\begin{tabular}{c|c|c|c}
    \hline \texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
  \end{tabular}
\end{small}
\caption{An execution on a queue separated by the operations in the root blocks.}
\end{table}

\noindent The \texttt{Dequeue} operations return \texttt{null, 5, 2, 1, 3, 4, null}, respectively.
Now, we claim that by knowing the size of the queue, we can compute the rank of the required \texttt{Enqueue} for any non-null \nf{Dequeue}. We apply this approach to blocks; if we store the size of the queue after each block of operations, we can compute the index of each \nf{Dequeue}'s result in \textsc{O}$(1)$ steps.

\begin{table}[hbtp]
\centering
\begin{small}
  \begin{tabular}{c|c|c|c|c}
    \hline &\texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
    \#\tt{Enq}s & 0 & 3 & 1 & 1 \\ \hline
    \#\tt{Deq}s & 1 & 1 & 1 & 4 \\ \hline
    Size at end & 0 & 2 & 2 & 0 \\ \hline
  \end{tabular}
\end{small}
  \caption{Augmented history of operation blocks on the queue.}
\end{table}

\noindent The size of the queue after the $b$th block in the root could be computed as
$$\textrm{max}\Big(\textrm{size after }(b-1)\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} - \#\textrm{\tt{Dequeue}s in }b\textrm{th block}, 0\Big).$$
\noindent Moreover, the total number of non-null dequeues in blocks $1,2,...,b$ in the root is
$$ \sum_{i=1}^{b} \#\textrm{\tt{Enqueue}s in }i\textrm{th block} - \textrm{size after }b\textrm{th block}.$$

Given a \texttt{Dequeue} is in block $B$, its response is the argument of the \tt{Enqueue} whose rank is
\begin{center}
    \#\textrm{non-null } \nf{Dequeue}s in  blocks $1,2,...,b-1 +  $index of the \nf{Dequeue} among B\textrm{'s} \nf{Dequeue}
\end{center}
if \vspace{-3em}
\begin{center}
    $\big ($size of the queue after $b-1\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} $\\$- \textrm{index of \tt{Dequeue} in }B\textrm{'s \tt{Dequeue}s} \big )\geq 0$.
\end{center}
Otherwise, the response would be \nf{null}.

\subsection{Details of the Implementation}
Section \ref{algQ} gives the pseudocode for the queue implementation. It uses the following two types of objects.
\paragraph{\tt{Node}} 
 In each \texttt{Node} we store pointers to its parent and its left and right child, an array of \nf{Block}s called \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.

\paragraph{\tt{Block}}
 The information stored in a \nf{Block} depends on whether  the \nf{Block} is in an internal node or a leaf. If it is in a leaf, we use a \nf{LeafBlock} which stores one operation. If a block $B$ is in an internal node $n$, then it contains subblocks in the left and right children of $n$. The left subblocks of $B$ are some consecutive blocks in the left child of $n$ starting from where the left subblocks of the block prior to $B$ ended. The right subblocks of $B$ are defined similarly. In each \texttt{block} we store four essential fields that implicitly summarize which operations are in the block \texttt{sum\textsubscript{enq-left}}, \texttt{sum\textsubscript{deq-left}}, \texttt{sum\textsubscript{enq-right}}, \texttt{sum\textsubscript{deq-right}}. The \texttt{sum\textsubscript{enq-left}} field is the total number of \nf{Enqueue} operations in the blocks before the last subblock of $B$ in the left child. The other fields' semantics are similar. The \nf{end\sub{left}} and \nf{end\sub{right}} field store the index of the last subblock of a block in the left and the right child, respectively. The approximate index of the superblock of non-root blocks is stored in their \nf{super} field. The \texttt{size} field in a block in the root node stores the size of the queue after the operations in the block have been performed.

We now describe the routines used in the implementation.

\paragraph{\tt{Enqueue($e$)}}
An \nf{Enqueue} operation does not return a response, so it is sufficient to propagate the \nf{Enqueue} operation to the root and then use its position in the linearization for future \nf{Dequeue} operations. \nf{Enqueue($e$)} creates a \nf{LeafBlock} with $\nf{element}=e$, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields and then appends it to the tree.

\paragraph{\tt{Dequeue()}}
\nf{Dequeue} creates a \nf{LeafBlock}, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields and appends it to the tree. Then, it computes the position of the appended \nf{Dequeue} operation in the root using \nf{IndexDequeue} and after that finds the response of the \nf{Dequeue} by calling \nf{FindResponse}.

\paragraph{\tt{Append($B$)}}
The \nf{head} field is the index of the first empty slot in \nf{blocks} in a \nf{LeafBlock}. There are no multiple write accesses on \nf{head} and \nf{blocks} in a leaf because only the process that the leaf belongs to appends to it. \nf{Append($B$)} adds $B$ to the end of the \nf{blocks} field in the leaf, increments \nf{head} and then calls \nf{Propagate} on the leaf's \nf{parent}. When \nf{Propagate} terminates, it is guaranteed that the appended block is a subblock of a block in the \nf{root}.

\paragraph{\tt{Propagate()}}
\nf{Propagate} on node $n$ uses the double refresh idea described earlier and invokes two \nf{Refresh}es on $n$ in Lines \ref{firstRefresh} and \ref{secondRefresh}. Then, it invokes \nf{Propagate} on \nf{$n$.parent} recursively until it reaches the root.

\paragraph{\tt{Refresh()} and \tt{Advance()}}
The goal of a \nf{Refresh} on node $n$ is to create a block of $n$'s children's new blocks and append it to $n$\nf{.blocks}. The variable \nf{h} is read from $n$\nf{.head} at Line \ref{readHead}. The new block created by \nf{Refresh} will be inserted into $n\nf{.blocks[h]}$. Lines \ref{startHelpChild1}--\ref{endHelpChild1} of \nf{$n$.Refresh} help to \nf{Advance} $n$'s children. \nf{Advance} increments the children's \nf{head} if necessary and sets the \nf{super} field of their most recently appended blocks. The reason behind this helping is explained later when we discuss \nf{IndexDequeue}. After helping to \nf{Advance} the children, a new block called \nf{new} is created in Line \ref{invokeCreateBlock}. Then, if \nf{new} is empty, \nf{Refresh} returns \nf{true} because there are no new operations to propagate, and it is unnecessary to add an empty block to the tree. Later we will use the fact that all blocks contain at least one operation. Line \ref{cas} tries to install \nf{new}. If it was successful, all is good. If not, it means someone else has already put a block in $n$\nf{.blocks[h]}. In this case, \nf{Refresh} helps advance $n$\nf{.head} to \nf{h+1} and update the \nf{super} field of $n$\nf{.blocks[h]} at Line \ref{advance}.

\paragraph{\tt{CreateBlock()}} \texttt{$n$.CreateBlock($h$)} is used by \nf{Refresh} to construct a block containing new operations of $n$'s children.
The block \nf{new} is created in Line \ref{initNewBlock} and its fields are filled similarly for both left and right directions. The variable  \nf{index\sub{prev}} is the index of the block preceding the first subblock in the child in direction \nf{dir} that is aggregated into \nf{new}. Field \texttt{new.end\textsubscript{dir}} stores the index of the rightmost subblock of \nf{new} in the child. Then \nf{sum\sub{enq-dir}} is computed from the sum of the number of \nf{Enqueue} operations in the \nf{new} block from direction \nf{dir} and the value stored in \texttt{$n$.blocks[h-1].sum\sub{enq-dir}}. The field \nf{sum\sub{deq-dir}} is computed similarly. Then, if \nf{new} is going to be installed in the \nf{root}, the \nf{size} field is also computed.

\paragraph{\tt{IndexDequeue($b,i$)}}

>>>>

% !TEX root =  podc-submission.tex

@@
